Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes: Overall people are guessing very low numbers of perceived audience

***

### Scatterplots
Notes:
```{r Scatterplots}
library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

qplot(x = age, y = friend_count, data=pf)

# Same thing, different code (it knows x comes first, y comes second)
qplot(age, friend_count, data= pf)
```

***

#### What are some things that you notice right away?
Response:
Younger people have a larger range for friend_count
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}

# Our previous code
qplot(x=age, y=friend_count, data=pf)

# Geom_point is a scatter plot, geom holds the type of graph
# Add one additional layer to debug plots (the plus signs)
ggplot(aes(x=age, y=friend_count), data=pf) + geom_point() +
  xlim(13, 90)

# Finding out lower and upper limits
summary(pf$age)
```

***

### Overplotting
Notes: 
Transparency of the points can be set with alpha operator in geom. (1/20 means that 20 points can only make up one dot in graph)

With geom_jitter we get a more dispersed distribution with a fuller understanding of the data compared to geom_point()

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_jitter(alpha = 1/20) +
  xlim(13,90)
```

#### What do you notice in the plot?
Response:
The time in your life that you have the most friends is in high school (age 14-18)
***

### Coord_trans()
Notes:

```{r Coord_trans()}
?coord_trans()

ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_point(alpha=1/20) + 
  xlim(13,90) + 
  coord_trans(y = "sqrt")

```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

geom_jitter is a convenient shortcut for a type of geom_point graph.
It adds a small amount of random variation to the location of each point and is a useful way of handling overplotting caused by discreteness in smaller datasets.

position = position_jitter() is used because for sqrt(0) a negative number is created and that is not a valid point for the scatter graph. position = position_jitter() removes these 0s 

```{r Alpha and Jitter}
?geom_jitter
summary(pf$age)

ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_jitter(alpha=1/10, position = position_jitter(h = 0)) +
  xlim(13,113) +
  coord_trans(y='sqrt')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:
dplyr package splits a dataframe and apply a function to some parts of the data. common functions are filter(), group_by(), mutate(), and arrange()

```{r Conditional Means}
install.packages('dplyr')
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())

pf.fc_by_age <- arrange(pf.fc_by_age, age)

head(pf.fc_by_age)

```

### Conditional Means Alternate Code
```{r}
install.packages('dplyr')
library(dplyr)

pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)

# same exact table!
head(pf.fc_by_age)
```


Create your plot!

```{r Conditional Means Plot}

ggplot(aes(x=age, y = friend_count_mean), 
       data = pf.fc_by_age) + geom_line()

```

***

### Overlaying Summaries with Raw Data
Notes:
To plot the summary with the data, we added the geom_line for the summary mean

?quantile()

?coord_cartesian()

The Cartesian coordinate system is the most familiar, and common, type of coordinate system. Setting limits on the coordinate system will zoom the plot (like you're looking at it with a magnifying glass), and will not change the underlying data like setting limits on a scale will.

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha=0.05, 
              position = position_jitter(h = 0),
              color = 'orange') +
  coord_cartesian(xlim = c(13, 90)) +
#  coord_trans(y = 'sqrt') + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs =  .5),
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9),
            color = 'blue', linetype = 2)
```

#### What are some of your observations of the plot?
Response: 
The mean and 0.5 summary line are not the same
More than 1000 friends is quite rare

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes: 

***

### Correlation
Notes:
Correlation coefficient measures the linear relationship between two variables

```{r Correlation}
?cor.test()

cor.test(pf$age, pf$friend_count, method = 'pearson')

with(pf, cor.test(age, friend_count, method = 'pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response: -0.027

***

### Correlation on Subsets
Notes:
Based on correlation and count, we've noticed that the relationship is not linear. We've also realized older ages are more likely to be incorrect. Take a look at 70 or less.

```{r Correlation on Subsets}
?subset()

with(subset(pf, age <= 70), cor.test(age, friend_count, method = 'pearson'))
```

***

### Correlation Methods
Notes:
We also have other measures of relationships than linear (pearson's r)

Monotonic relationships can be tested with different methods (like rho)

Single number coefficients are useful, but not a great substitute for scatter plots
***

## Create Scatterplots
Notes:


```{r}
summary(pf$www_likes_received)

ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  xlim(0, 129953) +
  geom_point()

```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
cor.test(pf$www_likes_received, pf$likes_received)

#with(subset(pf, www_likes_received <= 250), cor.test(www_likes_received,
#                                                     likes_received, method =
 #                                                      'pearson'))
```

Response:

***

### Moira on Correlation
Notes:

Scatter plots visually show the relationship, coefficients give you the evidence of it. Typically, when a regression or model is used, you must measure correlation first because regressions assume all variables are independent of one another.

***

### More Caution with Correlation
Notes:

Correlation coefficients can be deceptive if not careful. Plotting data is best way to understand it.

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)

data(Mitchell)
?Mitchell

summary(Mitchell)

ggplot(aes(x=Temp, y = Month), data = Mitchell) +
  geom_point()
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x=Month, y = Temp), data = Mitchell) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

0

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

cor.test(Mitchell$Month, Mitchell$Temp)

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

ggplot(aes(x=Month, y = Temp), data = Mitchell) +
  geom_point() +
  scale_x_continuous(breaks = seq(0,203, 12))

range(Mitchell$Month)

# ?scale_x_continuous()
# seq()
```

***

### A New Perspective

What do you notice?
Response: A cyclical pattern occurs

Watch the solution video and check out the Instructor Notes!
Notes:

# Overlaying each year's data on top of one another to see data in context
ggplot(aes(x=(Month%%12),y=Temp), data=Mitchell)+
  geom_point()

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

ggplot(aes(x=age, y=friend_count), data=pf.fc_by_age) +
  geom_line()

head(pf.fc_by_age, 10)
pf.fc_by_age[17:19,]

pf$age_with_months <- pf$age + (12-pf$dob_month)/12

```

***

### Age with Months Means

```{r Age with Months Means}

install.packages('dplyr')
library(dplyr)

age_groups <- group_by(pf, age_with_months)
pf.fc_by_age_months <- summarise(age_groups,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age_months, age_with_months)
head(pf.fc_by_age_months)

# Alternate code

library(dplyr)

pf.fc_by_age_months <- %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

```

Programming Assignment
```{r Programming Assignment}

age_with_months_groups <- group_by

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

ggplot(aes(x=age_with_months, y=friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
         geom_line()

```

***

### Smoothing Conditional Means
Notes: Comparing the age_with_months (above) to age graph (below), it can be seen that the age_with_months graph has a lot more noise in it.

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x=age_with_months,
                 y=friend_count_mean), 
             data = subset(pf.fc_by_age_months,
                           age_with_months < 71)) +
  geom_line() + geom_smooth()

p2 <- ggplot(aes(x=age, y=friend_count_mean), 
       data = subset(pf.fc_by_age, age < 71)) +
         geom_line() + geom_smooth()

library(gridExtra)
grid.arrange(p2, p1, ncol = 1)
```

***

### Which Plot to Choose?
Notes:
Exploratory data analysis is meant to do just that, no need to choose! When presenting, you will have to choose 1 or 2 graphs to communicate your findings.

***

### Analyzing Two Variables
Reflection:

http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat401/Notes/401-bivariate-slides.pdf

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

