Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes: Examining friend count & age with gender being the third variable. Use dplyr package and group_by, summarize, and other functions to do same thing as above

```{r Third Qualitative Variable}
# box plot to see mean difference
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) +
  geom_boxplot() + stat_summary(fun.y = mean, geom = 'point', shape = 4)

# line graph to compare over age data
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat =
              'summary', fun.y = median)

library(dplyr)

summary(pf)
?data_frame

pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = 
              mean(friend_count),
            median_friend_count =
            median(as.numeric(friend_count)),
            n = n()) %>%
  ungroup() %>%
  arrange(age)

head(pf.fc_by_age_gender)


```

***

### Plotting Conditional Summaries
Notes: 

```{r Plotting Conditional Summaries}

ggplot(aes(x = age, y = median_friend_count),
       data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))

```

***

### Thinking in Ratios
Notes:
It can help thinking of plots in relative terms - like ratios. Ex. how many more friends do women have on average compared to men.

***

### Wide and Long Format
Notes:
Long format = repeated common values and many rows
Wide format = shared common values in a table

It's helpful to use both long & wide format for vivewing the data.

Reshape2 will be the package we use to edit our data between long and wide
***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)

# dcast is used because we want result to be dataframe. dcast requires dataframe in first value and the formula to filter it.

pf.fc_by_age_gender_wide <-
  dcast(pf.fc_by_age_gender,
        age ~ gender,
        value.var = 'median_friend_count')

head(pf.fc_by_age_gender_wide)
```


***

### Ratio Plot
Notes:
Plot the rtion of the female to male median friend count using new dataframe created

```{r Ratio Plot}

?geom_hline

ggplot(aes(x = age, y = female/male), 
       data = pf.fc_by_age_gender_wide) +
  geom_line() + 
  geom_hline(yintercept = 1, alpha = 0.3,
             linetype = 2)

```

***

### Third Quantitative Variable
Notes:
Age & friend_count over a numerical variable

```{r Third Quantitative Variable}

summary(pf)

pf$year_joined <- floor(2014 - (pf$tenure/365))

summary(pf$year_joined)
table(pf$year_joined)
```

***

### Cut a Variable
Notes:
?cut 

cut buckets data on whatever variable you want

```{r Cut a Variable}

pf$year_joined.bucket <- cut(pf$year_joined,
                             c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:
Created a variable called year_joined and created year_joined_bucket as well. Let's see distribution


```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA = 'ifany')

ggplot(aes(x=age, y=friend_count), 
       data = subset(pf,!is.na(year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat =
                     'summary', fun.y = median)

```

***

### Plot the Grand Mean
Notes:
this is more of the high level exploration of the data that you want to be doing

```{r Plot the Grand Mean}

ggplot(aes(x=age, y=friend_count), 
       data = subset(pf,!is.na(year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat =
                     'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = mean, linetype =
              2)

```

***

### Friending Rate
Notes:

```{r Friending Rate}

sub_tenure <- subset(pf, tenure >= 1)
summary(sub_tenure$friend_count/sub_tenure$tenure)

```

***

### Friendships Initiated
Notes:
We've learned more about age & friend_count. Typically friends have higher friend counts if they have been on the site longer.

What is the median friend rate?
0.2205 (see above)
What is the maximum friend rate?
417 (see above)
```{r Friendships Initiated}

ggplot(aes(x = tenure, y = 
             friendships_initiated/tenure), data =
         subset(pf, tenure >= 1)) + 
         geom_line(aes(color = year_joined.bucket))

```

***

### Bias-Variance Tradeoff Revisited
Notes:

?geom_smooth

# friendships decline as tenure increases

ggplot(aes(x = tenure, y = 
             friendships_initiated/tenure), data =
         subset(pf, tenure >= 1)) + 
         geom_smooth(aes(color = year_joined.bucket))

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

Notices trends and then starts to trnsform it based on initial thoughts. Exoloratory Data Analysis is working through models and tranformations of the data to figure out what the best visualization and insight of the data actually is.

***

### Introducing the Yogurt Data Set
Notes:

Lots of yogurt data
-Over 2000 observations for 9 different variables
-For US households that buy yogurt over time

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}

getwd()

yo <- read.csv('yogurt.csv')

?str
str(yo)
summary(yo)

# Change id from an int to a factor
yo$id <- factor(yo$id)
str(yo)


library(ggplot2)
ggplot(aes(x = time, y = price), data = yo) +
  geom_histogram()

qplot(yo$price, geom = 'histogram')

```

***

### Number of Purchases
Notes:

distribution of yogurt is off

```{r Number of Purchases}
# 75th percentile is same as maximum
summary(yo)

?transform
yo <- transform(yo, all_purchases = mixed.berry + plain +
                             blueberry + strawberry + pina.colada)

```

***

### Prices over Time
Notes:

```{r Prices over Time}

qplot(x = all_purchases, data = yo, binwidth = 1, fill = I('#099009'))

ggplot(aes(x = time, y = price), data = yo) +
  geom_jitter(alpha = 1/4, shape = 21)

```

***

### Sampling Observations
Notes:

When familiarizing yourself with a dataset that hascontains multiple observations of the same units, it is often helpful to use sample of the dataset so it is easy to display the raw data for that sample. IN case of yoghurt, we may want to look at households in more detail.

To first understand this dataset, first we would ant to find out how often households buy multiple items, yoghurt, and what prices they are buying at.

Picking 16 households at random to take a closer look and better understand these points.

***

### Looking at Samples of Households

# Note: x %in% y returns a logical (boolean) vector the same length as x that says whether each entry in x appears in y. That is, for each entry in x, it checks to see whether it is in y. This allows us to subset the data so we get all the purchases occasions for the households in the sample. Then, we create scatterplots of price vs. time and facet by the sample id. 

```{r Looking at Sample of Households}

set.seed(4230)

sample.ids <- sample(levels(yo$id), 16)

# size of point makes consideration for each scatter plot

ggplot(aes(x = time, y = price),
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id) +
  geom_line() +
  geom_point(aes(size = all_purchases), pch = 1)

```

***

### The Limits of Cross Sectional Data
Notes:

If we have observations over time, we can look at the individual dataset (house for yoghurt data).

The facebook data is not great for friendship over time, but if we had a daataset like the yogurt's we would be able to compare friendship over time.

***

### Many Variables
Notes:

We use EDA to let the variables/relationships/questiion of interest to speak and surprise us. We look to incorporate more variables into a relationship to give us more information.

Visualizations and summaries are helpful to tell us more about the dataset and variables of interest.

We can help speed up our EDA by creating many plots or comparisons at once. This is one way to let the dataset speak to us, especially for variables we may not have been examining.

***

### Scatterplot Matrix
Notes:

# Scatterplot Matrix creates a bunch of plots all at once (a scatter plot of all paired variables). Categorical variables 

install.packages('GGally')
library(GGally)
theme_set(theme_minimal(20))

getwd()
setwd("Desktop/education/udacity/data_analyst_nanodegree/project_4/lesson 5")

pf <- read.table("pseudo_facebook.tsv")

# set the seed for reproducible results

set.seed(1836)
names(pf)
names(pf_subset)
pf_subset <- pf[, c(2:15)]
names(pf_subset)

# this will take a long time to plot!
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])

# this should take less time
# pf_subset = pf[, c('age', 'dob_year', 'dob_month', 'gender', 'tenure')]
pf_subset_2 <- subset(pf, select = -c(userid, year_joined, year_joined_bucket))

***

### Even More Variables
Notes:

# Genomic data has thousands of data points to have to look through, hence matrix scatterplots are so important.

getwd()

nci <- read.table('nci.tsv')

colnames(nci) <- c(1:64)

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

# finding out when a gene is overexpressed in red and underexpressed in blue

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

