Analysis of Red Wine by Marc Maybaum
========================================================

Studying oenology in college for a semester, I've become knowledgable in wine,
but I thought it would be interesting to have the perspective of wine data to 
better understand wine in a different way.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all packages used
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(PerformanceAnalytics)
library(corrplot)
library(RColorBrewer)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# getwd()

redwine <- read.csv('wineQualityReds.csv', row.names = NULL, header = TRUE)
summary(redwine)
```

The dataset consists of 1599 observations and 13 variables. Because the X
variable is just a list of indexes, I will be removing it from the analysis.

```{r echo=FALSE, Remove_X}
# getwd()

redwine$X <- NULL
```

# Univariate Plots Section

### Alcohol Histogram

The average alcohol percentage is 10.42%. The majority of wines are above
8% and less than 14%.

```{r echo=FALSE, Univariate_Plot_Alcohol}

# Creating a function for repetitive code building out histograms
# I am keeping all previous code in file in case of later use

create_plot <- function(varname, binwidth = 0.1) {
  return(ggplot(aes_string(x = varname), data = redwine) 
          + geom_histogram(binwidth = binwidth))
  }

create_plot('alcohol', 0.1) + scale_x_continuous(lim = c(8, 15))

# ggplot(data = redwine, aes(x = alcohol)) + 
#  scale_x_continuous(lim = c(8, 15)) +
# geom_histogram(binwidth = 0.025)
```

### Density Histogram

Density has an even distribution with wines having a mean density of 0.9967.

```{r echo=FALSE, Univariate_Plot_Density}

create_plot('density', 0.00025)

# ggplot(data = redwine, aes(x = density)) +
#  geom_histogram(binwidth = 0.00025)

```


### pH Histogram

pHs have little variation with a minimum value of 2.74 and a maximum of 4.01.

```{r echo=FALSE, Univariate_Plot_pH}

create_plot('pH', 0.01) + scale_x_continuous(lim = c(2.5, 4.2))

# ggplot(data = redwine, aes(x = pH)) + 
#  scale_x_continuous(lim = c(2.5, 4.2)) +
#  geom_histogram(binwidth = 0.01)
```

### Total Sulfur Dioxide Histogram

Total Sulfur Dioxide is positively skewed. To see if there is a clearer
picture of the measurement, I am going to visualize the log of it. Total Sulfur Dioxide has more of a normal distribution after taking the log
of the measurement.

```{r echo=FALSE, Univariate_Plot_TotalSulfur}

plotA <- ggplot(data = redwine, aes(x = total.sulfur.dioxide)) + 
  geom_histogram(binwidth = 2.5)

plotB <- ggplot(data = redwine, aes(x = log(total.sulfur.dioxide))) +
  geom_histogram(binwidth = 0.05)

grid.arrange(plotA, plotB, ncol = 2)

```


### Sulphates Histogram

The majority of sulphates are bunched between 0 and 1.

```{r echo=FALSE, message = FALSE, warning=FALSE, Univariate_Plot_Sulphates}

create_plot('sulphates', 0.01) + scale_x_continuous(lim = c(0.25, 1.5))

# ggplot(data = redwine, aes(x = sulphates)) +
#   scale_x_continuous(lim = c(0.25, 1.5)) +
#  geom_histogram(binwidth = 0.01)

```

### Residual Sugar Histogram

The histogram of residual sugar was positively skewed. Upon taking the log
of residual sugar, we get a distribution closer to a normal distribution with
a tail at the end of the x axis.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_Sugar_Raw}

plot1 <- ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.1)

plot2 <- ggplot(data = redwine, aes(x = log(residual.sugar))) +
  geom_histogram(binwidth = 0.05)

grid.arrange(plot1, plot2, ncol = 2)
```

Zooming in to view the normal distribution between 0.25 & 4, we can see that
it is likely for the residual sugars to fall between 0 and 4, though there are
other wines with higher concentrations of residual sugars (ports and dessert
wines).

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_Sugar_Zoomed}

create_plot('residual.sugar', 0.1) + scale_x_continuous(lim = c(0.25, 4))

# ggplot(data = redwine, aes(x = residual.sugar)) +
#   scale_x_continuous(lim = c(0.25, 4)) +
#   geom_histogram(binwidth = 0.05)
```


### Quality Histogram

Quality is an ordinal variable. Because of this I used a bar chart to measure
the count of each category (quality score).

```{r echo=FALSE, Univariate_Plot_Quality}

ggplot(data = redwine, aes(x = as.factor(quality))) + 
  geom_bar()

```

```{r echo=FALSE, Univariate_Summary_6}
summary(redwine$quality)
```

# Univariate Analysis

### What is the structure of your dataset?

Red Wine dataset is made up of 1599 observations and 12 variables (Fixed Acidity, Volatile Acidity, Citric Acid, Residual Sugar, Chlorides, Free Sulfur Dioxide, Total Sulfur Dioxide, Density, pH, Sulphates, Alcohol, Quality).
Observations thus far:

- Alcohol levels vary between 8% and 14%
- Many wines have similar pH levels (between 2.74 & 4.01)
- The majority of wines have less than 1.0 sulphates
- The quality for wines vary, but the median rating is 6 

### What is/are the main feature(s) of interest in your dataset?

My main interest in the dataset is quality of wine. If any of the variables will have a relationship with quality of wine, I think it would be acidity or sulphates.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I think it will be rare to find a strong relationship between many of the measurements (alcohol %, pH, sulphates, residual sugars, and acidity) and quality. My hunch is that acidity will have the strongest relationship with quality score.

### Did you create any new variables from existing variables in the dataset?

No, but I did remove an unnessarcy variable for analysis (X).

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Both Total Sulfur Dioxide and Residual Sugars had a positive skew 
visualization. Upon transforming the data using the log() formula, the
distribution of the data became more normally distributed appearing. I did this
to see if I could have a better understanding of the data without having the
outlying data impact the visualization seen.

# Bivariate Plots Section

First, I want to calculate the Pearson Correlation Coefficient to find
out if there is any relationship between the variables. I will first create a
table of all these calculations and then turn the table into multiple
visualizations.

```{r echo=FALSE, Bivariate_Plots_Matrix}
# correlation table using Pearson r value

wine_cor <- cor(redwine, method = "pearson")
wine_cor
```

The below is a heatmap of the different variables and how correlated they are
to one another (seen above for actual scores). When the variable is compared to 
itself, the correlation is a perfect 1. This design can be seen diagonally in 
the visual.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots_Heatmap}
# help from http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software

# heatmap to showcase strength of relationships between all variables

col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = wine_cor, col = col, symm = TRUE)
```


Of the relationships, the strongest positive correlations are between acids 
and sulfur measurements. Citric acid & fixed acid and free sulfur dioxide & 
total sulfur dioxide are the strongest positively correlated variables. 
Additionally, of the negatively correlated relationships, pH & citric acid and
pH & fixed acidity are very strong relationships.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_Table}
# install.packages("corrplot")

# creating a plot of correlations using size and opaqueness to symbolize
# correlation coefficient

corrplot(wine_cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

Further looking into these relationships in the above visualization, it is
easier to view the stronger relationships based on the size and opaqueness 
of the circle. Some additional relationships that were not identified previously
include negative relationships between alchol & density and volatile acidity & 
citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}

# install.packages('PerformanceAnalytics')

# creating correlation table with histograms between every variable

chart.Correlation(redwine, histogram=TRUE, method = 'pearson')

```

Lastly, I've created a correlation chart that is able to showcase the
correlation coefficients with a scatterplot of the two variables. We can see
many of the stronger correlations identified earlier in the other 
visualizations.


```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Plots_Alcohol}

# becuase quality is an ordinal varibale, we must use factor() to properly
# visualize the variable
quality_wine <- factor(redwine$quality) 

ggplot(redwine, aes(x = quality_wine, y = alcohol)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), alpha = 0.25) +
  labs(title = "Quality & Alcohol's Dispersion")

#boxplot(alcohol~quality, data=redwine, main="Quality & Alcohol's Relationship",
#        xlab = 'Quality', ylab = 'Alcohol %') +
#  geom_jitter(position=position_jitter(width=.1, height=0))  
```

Looking into the strongest relationships for red wine variables, I've created 
box plots, which are able to show the mean, quartile, and outlying variables for
each quality score. Additionally, I've added jitter plots to the boxplots, which
shows the quantity of data dispersed within each score. Alcohol & Quality can be
seen above and Volatile Acidity & Quality can be seen below.


```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Plots_VolatileAcidity}

ggplot(redwine, aes(x = quality_wine, y = volatile.acidity)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), alpha = 0.25) +
  labs(title = "Quality & Volatile Acidity's Dispersion")

#ggplot(aes(x = quality, y = volatile.acidity),
#       data = redwine) + geom_point(alpha = 0.25, size = 0.5) + 
#  stat_summary(fun.y = mean, geom = 'point', color = 'blue', 
#               size = 3)
```

# Bivariate Analysis

### How did the feature(s) of interest vary with other features in the dataset?

Quality showed very weak relationships between many variables in the dataset.
Of the strongest relationships, Alcohol & Quality showed a moderate positive
relationship (r^2 value of 0.48) and Volatile Acidity & Quality had a moderate
negative relationship (r^2 value of -0.39). These two observations are
interesting because it gives clues to potential requirements of what a wine 
taster expects from a quality wine - a threshold of alcohol (possibly to balance
the wine profile) and lower levels of Volatile Acidity (possible expectation for 
how much acid should be contained in a wine). Because the Pearson value is only 
a moderate relationship for both of these relationships, this also implies that 
these expectations are not consistently held between tasters compared to if a 
strong relationship would exist. Both of these observations can be seen in the
above bivariate plots - mean Alcohol level increases as Quality increases and
mean Voltaile Acidity decreases as Quality of Wine increases.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

All of the strongest relationships came from variables other than my main 
focus of interest. In fact, quality of wine (as expected) is not strongly 
correlated with any of the variables supplied in this dataset. My strongest
relationships were between pH & Fixed Acidity (r^2 value of -0.68), Fixed
Acidity & Density (r^2 value of 0.67), Fixed Acidity & Citric Acid (r^2 value of
0.67), and Free Sulfur Dioxide & Total Sulfur Dioxide (r^2 value of 0.67).

### What was the strongest relationship you found?

The strongest relationship was between pH and Fixed Acidity at -0.68

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_LowQuality}

# shows common low quality wine scatter plot coordinates
ggplot(redwine, aes(x = alcohol, y = volatile.acidity, color = quality_wine)) +
  scale_x_continuous(lim = c(8.5,14)) +
  scale_y_continuous(lim = c(0,1.7)) +
  scale_color_brewer(type = 'seq', palette = "Blues", direction = -1) +
  theme_dark() +
  geom_point(alpha = 0.4, size = 1, position = 'jitter') +
  ggtitle("Alcohol & Volatile Acidity Levels by Quality of Wine")
```

When graphing Voltatile Acidity & Alcohol by Quality of Wine, it is noticeable
that the lowest quality wines are grouped together in the lower alcohol level of
the graph with a higher variation and higher levels of volatile acidity, and the
higher quality wines are grouped together in the higher levels of alcohol
percentage and lower levels of volatile acidity. This visualization is also in 
line with the correlation levels between each variable and quality of wine found
in the bivariate plots section.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_HighQuality}

# Shows common high quality wine scatter plot coordinates with alcohol & 
# volatile acidity taken into consideration

ggplot(redwine, aes(x = alcohol, y = volatile.acidity, color = quality_wine)) +
  scale_x_continuous(lim = c(8.5,14)) +
  scale_y_continuous(lim = c(0,1.5)) +
  scale_color_brewer(type = 'div', palette = "RdYlGn") +
  theme_dark() +
  geom_point(alpha = 0.4, size = 1, position = 'jitter') +
  ggtitle("Alcohol & Volatile Acidity Levels by Quality of Wine")
```

Graphing the same variables with a different color variation shows the same
observation previously identified - high quality wines are seen, generally, in 
the lower, right hand corner of the graph with green points. It is noticeable 
that higher alcohol and lower and less varied volatile acidity 
levels are the common themes for higher quality wines, though there are few of 
these wines compared to the rest of the dataset (see below).


```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Table}
print("Count of Quality Score")
table(redwine$quality)
```

After understanding the relationship between Quality & Alcohol and Quality &
Volatile Acidity, let's see how well we can create and train a model with the 
top 5 correlated variables for quality of wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Model}
m1 <- lm((I(quality) ~ I(alcohol)), data = redwine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
mtable(m1, m2, m3, m4, m5)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Model_2}
summary(m5)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Model_3}
new_wine = data.frame(alcohol = 12.3, volatile.acidity = 0.34,
                      citric.acid = 0.17, sulphates=0.6581, 
                      total.sulfur.dioxide = 35.43)

modelEstimate = predict(m5, newdata = new_wine,
                        interval="prediction", level = .95)
exp(modelEstimate)
```

As can be seen in the r-squared output & the model's prediction for a quality
score with an example dataset, the relationships between the variables in the 
red_wine dataset are not strong enough to predict the quality score of a wine.
Even with the 5 strongest relationships accounted for, we are only able to reach
a 0.344 r-squared value for the model. Unfortunately, this model needs stronger
relationships and a larger data set to be trained to give an accurate
prediction. This can also be seen in the below residual plot of our model.


```{r echo=FALSE, message=FALSE, warning=FALSE, Residual_Plot_Model}

redwine_resid <- resid(m5)

plot(redwine$quality, redwine_resid, ylab = "Residuals", xlab = "Quality Score", main = "Residual Values of Each Quality Score")

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The moderate linear relationships between quality & alcohol/volatile acid
showcased the pattern that each individual relationship had on quality. Although
because the relationships were moderate, sometimes these visualizations were
harder to showcase. High quality wines in particular had a general pattern, but
needed a lot more data to substantiate visual representation.

### Were there any interesting or surprising interactions between features?

After viewing the correlations in multiple visualizations, it was pretty obvious
how variables interacted with one another. If anything, the multivariate
analysis showcased the weakness of a lot of the variables' relationship with
quality score.

### Did you create any models with your dataset? Discuss the strengths and\
limitations of your model.

I created a model with the weak relationships of the variables to quality.
As expected, the model had a weak coefficient to predicting future results with 
only an r-squared value of 0.344. The prediction output wasn't even within the 
range of quality scores in the dataset. Much more data and stronger 
relationships would help the model gain a higher r-squared value and a more
accurate expected output.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(redwine, aes(x = quality_wine, y = alcohol)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), alpha = 0.25) +
  labs(title = "Quality & Alcohol's Relationship",
       x = "Quality (score between 0 and 10)",
       y = "Alcohol (% by volume)")
```

### Description One

As can be seen in this box plot, alcohol and quality are plotted against
one another. Additionally, the mean amount of alcohol in each quality score is
visualized with the line in the box. As shown from the points within each
quality score, 5 & 6 scores contain the majority of the red_wine data.


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(redwine, aes(x = alcohol, y = volatile.acidity, color = quality_wine)) +
  scale_x_continuous(lim = c(8.5,14)) +
  scale_y_continuous(lim = c(0,1.5)) +
  scale_color_brewer(type = 'div', palette = "RdYlGn") +
  theme_dark() +
  geom_point(alpha = 0.4, size = 1, position = 'jitter') +
  labs(x = "Alcohol (% by volume)", y = "Volatile Acidity (g/dm^3)",
       color = "Quality (score between 0 and 10)", 
       title ="Alcohol & Volatile Acidity Levels by Quality of Wine")
```

### Description Two

The scatter plot of volatile acidity & alcohol colored by the quality score 
shows the common traits of a quality wine based on these two traits. Of the low quality scores, wines have a lower alcohol percentage and a higher amount of volatile acidity than highly scored wines. The low quality wines can be seen in 
the upper left portion of the scatter plot in yellow, while the high quality 
wines are not as visible in the lower, right area of the scatter plot in green.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE, results=FALSE, Plot_Three}

plot(redwine$quality, redwine_resid, ylab = "Standardized Residual Value", 
     xlab = "Quality (score between 0 and 10)", 
     main = "Residual Plot of Each Quality Score")
```

### Description Three

The residual plot is made from the model created from the 5 strongest 
correlated variables (m5) to quality score. The residual is a standardized value
that is computed by taking the difference between the actual and predicted
values of Quality Score in our Red Wine dataset. As can be seen in the graph,
Quality Scores 5 and 6 are closest to averaging a 0 Standardized Residual Value.
This means that these two Quality Scores have the best predicted scores
according to our model. These are also the scores comprising the majority of 
data in our red wine dataset.


# Reflections

After exploring the Red Wine dataset I have learned about wine in ways I did
not expect. The biggest lesson is that it is very tough to know how a wine will 
be received by the consumer based on pure data measurements. Though there are
some common variables for a higher quality score, like a low amount of volatile 
acidity and at least a mean amount of alcohol, there is still much that could be 
learned from a larger data set and measurements. I think I was cautious enough 
when exploring the data to not make many assumptions and dug into the correct 
data available to analyze quality score. 

My difficulties were most significant when trying to develop a model based on 
the available data. There were both not enough significant relationships and 
enough data to train the model to accurately predict quality score ratings. In 
the future, enriching the data with more collection and new measurements could 
possibly help train the model to predict quality score.

